D3064R0: Bringing the Laws of Physics to Bear on the OOTA Problem

### 2024-01-10 ootavacuous.tex
Add appendix showing UB-induced OOTA cycles
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
either zero or one, if the values returned by the two loads from \co{x}
differ, then one of those two values must be equal to that from \co{y}.
If the implementation is further permitted to make a runtime choice
between the two values loaded from \co{x}, it could always choose the
one that was equal to the value loaded from \co{y}, as shown by the
\co{if} statement on line~13.
This would mean that the store to \co{z} was always the constant one,
which would permit the store to be ordered before the load from \co{y},
as shown in lines~14 and~15.
Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.

@@

>>>>

Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
either zero or one, if the values returned by the two loads from \co{x}
differ, then one of those two values must be equal to that from \co{y}.
If the implementation is further permitted to make a runtime choice
between the two values loaded from \co{x}, it could always choose the
one that was equal to the value loaded from \co{y}, as shown by the
\co{if} statement on line~13.
This would mean that the store to \co{z} was always the constant 1,
which would permit the store to be ordered before the load from \co{y},
as shown in lines~14 and~15.
Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.

@@

\subsubsection{OOTA From Undefined-Behavior Optimizations}
\label{app:OOTA From Undefined-Behavior Optimizations}

Many C++ implementations carry out optimizations based on the assumption
that UB cannot happen, and these optimizations can result in OOTA cycles,
with a simple example shown in
Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-div-ub.litmus @@
\caption{OOTA Due To Divide-By-Zero UB}
\label{lst:OOTA Due To Divide-By-Zero UB}
\end{listing}

On line~14 of this listing, the value of \co{r2} must be either zero
or one due to the comparison on line~13.
If it is the value one, then there is a divide-by-zero error, which
enables the C++ implementation to assume that the value of \co{r1} is
always zero.
This in turn allows the implementation to behave as if the expression
stored to \co{y} on line~14 was the constant 1, at which point there
is no semantic dependency from the load from \co{x} on line~13 to the
store to \co{y} on line~14.

The implementation is then within its rights to act as if lines~13 and~14
were interchanged.
At which point, the order of execution could be as follows:

\begin{enumerate}
\item	Line~14 stores the constant 1 to \co{y}.
\item	Line~8 loads the constant 1 from \co{y}, setting \co{r1} to 1.
\item	Line~9 stores the constant 1 to \co{x}.
\item	Line~8 loads the constant 1 from \co{x}, setting \co{r2} to 1.
\end{enumerate}

Please note that the OOTA-like cycle called out in the \co{exists}
clause can happen even though the program never actually executes any
code that would have resulted in division-by-zero UB.
It is tempting to consider using the techniques described in
Appendix~\ref{app:Aside on Undefined Behavior}
that prevent back-propagation of UB, but these are ineffective in this
case because the C++ implementation can detect and exploit UB when
considering line~14 in isolation.
Once the constant 1 is substituted for line~14's expression, the
hardware is free to reorder lines~13 and~14 even if a compiler-based
C++ implementation refrains from doing so.

Therefore, this paper focuses on UB-free programs, rare though these
beasts might be among large-scale C++ programs running in production.
